---
title: "Timeseries Analysis"
author: "Nichols, PK"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

#1. Setup

## A. Load Packages

```{r setup, echo=TRUE, message=FALSE, warning=FALSE, results='hold'}
## ---- libraries ----
library(ggplot2)
library(tidyverse)
library(vegan)
library(iNEXT)
library(ggpubr)
library(codyn)
library(synchrony)
library(randomForest)
library(caret)
library(scales)
library(forcats)
library(zetadiv)
library(purrr)
library(tidyr)
library(caret)
library(ggspatial)
library(sf)
library(rnaturalearth)
library(zoo)
library(glmmTMB)
library(dplyr)
library(tidyr)
library(ggeffects)
library(circular)
library(mvtnorm)
library(MuMIn)
library(rfPermute)
library(parallel)
library(mgcv)
library(glmmTMB)
library(mvtnorm)
library(performance)
library(gratia)
library(patchwork)

## ---- reproducibility ----
set.seed(123)
available_cores=detectCores(logical=FALSE)  # physical cores
cores_to_use=max(1, available_cores - 1)

numperm=9999

## ---- paths ----
paths=list(
  data_raw      ="data/raw",
  data_processed="data/processed",
  results_rds   ="results/rds",
  results_tables="results/tables",
  results_figs  ="results/figures"
)

walk(paths, dir.create, showWarnings=FALSE, recursive=TRUE)

## ---- plotting theme ----
My_Theme=theme_set(theme_bw()+theme(
  panel.grid.major=element_blank(), 
  panel.grid.minor=element_blank(),
  panel.background=element_blank(), 
  panel.border=element_rect(fill=NA, colour="black"), 
  panel.spacing=unit(1, "lines"),
  axis.title.x=element_text(size=12, color="black"),
  axis.text.x=element_text(size=12, color="black"),
  axis.title.y=element_text(size=12, color="black"),
  axis.text.y=element_text(size=12, color="black"),
  legend.text=element_text(size=12, color="black"),
  strip.text.x=element_text(size=14, color="black"),
  strip.text.y=element_text(size=14, color="black"),
  strip.text=element_text(size=14, color="black"),
  strip.background=element_blank(), 
  strip.placement="outside",
  title=element_text(size=14, color="black", face="bold"),
  axis.title=element_text(size=12, face="bold", colour="black"), 
  axis.ticks=element_blank(), 
  axis.text=element_blank()))

My_Theme=theme_set(theme_bw()+theme(
  panel.grid.major=element_blank(), 
  panel.grid.minor=element_blank(),
  panel.background=element_blank(), 
  panel.border=element_rect(fill=NA, colour="black"), 
  panel.spacing=unit(1, "lines"),
  axis.title.x=element_text(size=12, color="black"),
  axis.text.x=element_text(size=12, color="black"),
  axis.title.y=element_text(size=12, color="black"),
  axis.text.y=element_text(size=12, color="black"),
  legend.text=element_text(size=12, color="black"),
  strip.text.x=element_text(size=14, color="black"),
  strip.text.y=element_text(size=14, color="black"),
  strip.text=element_text(size=14, color="black"),
  strip.background=element_blank(), 
  strip.placement="outside",
  title=element_text(size=14, color="black", face="bold"),
  axis.title=element_text(size=12, face="bold", colour="black"), 
  axis.ticks=element_blank(), 
  axis.text=element_blank()))

cols.site=c("#425754", "#DEB3AD","#A65B24", "#CD9629")
cols.marker=c("#a6cdc6", "#233e34")
cols.site.paired=c("#425754", "#83BEC6", "#D68C96", "#DEB3AD", "#A65B24", "#E4C188", "#CD9629", "#F4DF9D")

st_err=function(x) sd(x, na.rm=TRUE) / sqrt(sum(!is.na(x)))
```

# 2. Figures

## A. Figure 1

### i. Map data

```{r}
metadata=read.csv(file.path(paths$data_raw, "metadata_SST.csv"), header=T)

# Get the world dataset
world=ne_states(country="united states of america", returnclass="sf")

# Filter the dataset to focus on California
hawaii=world[world$name == "Hawaii", ]

# Define the locations and labels
locations=data.frame(
  Location=c("Maunalua", "Waimānalo", "Kāneʻohe", "Lanikai"),
  Lat=c(21.275176, 21.326470, 21.434942, 21.393037),
  Lon=c(-157.760664, -157.682236, -157.787055, -157.714236)
)
```

### ii. Panel A

```{r}

# Plot the map focusing on Oahu with colored points and labels
main_plot=ggplot(data=hawaii) +
  geom_sf() +
  # Drop shadow layer for points
  #geom_point(data=locations, aes(x=Lon, y=Lat),size=5, color="black") +
  geom_point(data=locations, aes(x=Lon, y=Lat, color=Location), size=4) +  
  #geom_text(data=locations, aes(x=Lon, y=Lat, label=Location),hjust=0.21, vjust=-0.6, size=6, fontface="bold", color="black", stroke=0.5) + # Black outline for labels
 geom_text(data=locations, aes(x=Lon, y=Lat, label=Location, color=Location), hjust=0.2, vjust=-0.6, size=6, fontface="bold") + # Main text labels
  annotation_scale(location="bl", width_hint=0.5) +
  annotation_north_arrow(location="bl", which_north="true", 
                         pad_x=unit(0.75, "in"), pad_y=unit(0.5, "in"),
                         style=north_arrow_fancy_orienteering) +
  scale_color_manual(values=cols.site) +
  labs(color="Location") +
  coord_sf(xlim=c(-158, -157.6), ylim=c(21.2, 21.5)) +
  My_Theme + 
  theme(legend.position="none") + labs(x="", y="")




# Data for the inset map (assuming inset_map is an sf object)
# Replace with your actual data
inset_map=hawaii

# Create the inset map plot with a smaller size and a box around it
inset_plot=ggplot(data=inset_map) +
  geom_sf() +
  geom_point(data=locations[1, ], aes(x=Lon, y=Lat), color="black", size=2) + # Marker for Honolulu
  coord_sf(xlim=c(-160, -154), ylim=c(18, 23)) + # Adjust the limits for the inset map
  theme_void() + # Remove axis labels and ticks
  theme(plot.background=element_rect(color="black", size=1)) # Add a black border around the plot

map=main_plot + annotation_custom(ggplotGrob(inset_plot), xmin=-158, xmax=-157.9, ymin=21.4, ymax=21.5) + 
  geom_text(aes(x=-157.985, y=21.445, label="Hawaiʻi"), 
            hjust=0, vjust=1, size=4, fontface="bold") +
  geom_text(aes(x=-157.985, y=21.45, label="Oʻahu"), 
            hjust=0, vjust=0, size=4, fontface="bold") + 
  geom_text(aes(x=-157.89, y=21.33, label="Honolulu"), 
            hjust=0, vjust=1, size=4, fontface="bold") + geom_point(aes(x=-157.8583, y=21.3069), color="black", size=3) +geom_point(aes(x=-157.8583, y=21.3069), shape=5, size=6) + theme(panel.grid=element_blank()) 

map
```

### ii. NMDS and Envfit data

```{r}
# ---- NMDS data ----
nmds.data.18S=read.csv(file.path(paths$results_tables,"NMDS_data_18S.csv"), header=T)
nmds.data.18S$MARKER="18S"

nmds.data.COI=read.csv(file.path(paths$results_tables,"NMDS_data_COI.csv"), header=T)
nmds.data.COI$MARKER="COI"

nmds.data=rbind(nmds.data.18S, nmds.data.COI)
```

### iii. Panel B

```{r}
nmds.site=ggplot(nmds.data, aes(x=NMDS1, y=NMDS2)) +
  geom_point(aes(colour=SITE), alpha=0.33) +
  geom_vline(xintercept=0, lty=2, col="grey66") +
  geom_hline(yintercept=0, lty=2, col="grey66") +
  stat_ellipse(aes(fill=SITE, colour=SITE),
               level=0.95, geom="polygon", alpha=0.3) +
  scale_fill_manual(values=cols.site) +
  scale_color_manual(values=cols.site) +
  facet_wrap(~ MARKER, scales="free") +
  My_Theme +
  theme(legend.position="right") 

nmds.site
```

###iv. Export

```{r}
fig1=ggarrange(
  map,
  nmds.site,
  ncol=1,
  nrow=2,
  heights=c(1.3, 1),
  widths=c(2, 1),
  labels="AUTO"
)

ggsave(
  filename=file.path(paths$results_figs, "FIGURE_1.pdf"),
  plot    =fig1,
  width   =6,
  height  =8,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

## B. Figure 2

### i. Zeta diversity data

```{r}
## Marker Decay
zeta.str.decay.18S=read.csv(file.path(paths$results_tables, "zeta_decay_scaled_marker18S.csv"), header=T)
zeta.str.decay.COI=read.csv(file.path(paths$results_tables, "zeta_decay_scaled_markerCOI.csv"), header=T)
zeta.str.decay=rbind(zeta.str.decay.18S, zeta.str.decay.COI)

## Marker Ratio
zeta.str.ratio.18S=read.csv(file.path(paths$results_tables, "zeta_ratio_marker18S.csv"), header=T)
zeta.str.ratio.COI=read.csv(file.path(paths$results_tables, "zeta_ratio_markerCOI.csv"), header=T)
zeta.str.ratio=rbind(zeta.str.ratio.18S, zeta.str.ratio.COI)

## Site-wise Decay
zeta.str.site.18S=read.csv(file.path(paths$results_tables, "zeta_decay_18S.csv"), header=T)
zeta.str.site.COI=read.csv(file.path(paths$results_tables, "zeta_decay_COI.csv"), header=T)
zeta.str.site=rbind(zeta.str.site.18S, zeta.str.site.COI)


#combined structure decay and ratio
zeta.str.decay$variable="Zeta Decay"
zeta.decay.long=zeta.str.decay %>%
  transmute(
    zeta.order,
    y=zeta.val,
    sd=zeta.val.sd,
    Marker,
    variable
  )

zeta.str.ratio$variable="Zeta Ratio"
zeta.ratio.long=zeta.str.ratio %>%
  transmute(
    zeta.order,
    y=ratio,
    sd=NA_real_,
    Marker,
    variable
  )

zeta.str=bind_rows(zeta.decay.long, zeta.ratio.long)
```

###ii. Decay and ratio plots

#### a. Panels A-B

```{r}
zeta.markerplot=zeta.str %>% 
  ggplot(aes(x=zeta.order, y=y, color=Marker)) + # Map color to MARKER
  geom_point() + # Points inherit color from aes
  geom_line() +  # Lines inherit color from aes
  geom_ribbon(
    aes(ymin=y - sd, ymax=y + sd, fill=Marker), 
    alpha=0.1, # Transparency for the ribbon
    color=NA # No border color
  ) +
  scale_color_manual(values=cols.marker) + # Custom colors for MARKER
  scale_fill_manual(values=cols.marker) +  # Ribbon fill colors based on MARKER
  ylab("Proportion") +
  xlab("") + facet_wrap(~variable, scales="free_y") +
  My_Theme + theme(legend.position="right") + xlim(0,16)+scale_y_log10() + ylim(0,1)
zeta.markerplot
```


#### b. Panels C-D

```{r}
zeta.siteplot=zeta.str.site %>% 
  ggplot(aes(x=zeta.order, y=zeta.val, color=Site)) + # Map color to MARKER
  geom_point() + # Points inherit color from aes
  geom_line() +  # Lines inherit color from aes
  geom_ribbon(
    aes(ymin=zeta.val - zeta.val.sd, ymax=zeta.val + zeta.val.sd, fill=Site), 
    alpha=0.1, # Transparency for the ribbon
    color=NA # No border color
  ) +
  scale_color_manual(values=cols.site) + # Custom colors for MARKER
  scale_fill_manual(values=cols.site) +  # Ribbon fill colors based on MARKER
  ylab("Proportion") +
  xlab("Zeta Order") + facet_wrap(~Marker) +
  My_Theme + theme(legend.position="right")+ xlim(0,16) + ylim(0,1)+ scale_y_log10()
```

###iii. Export

```{r}
fig2=ggarrange(zeta.markerplot, zeta.siteplot, ncol=1, nrow=2, align="hv", legend="none")
fig2
ggsave(
  filename=file.path(paths$results_figs, "FIGURE_2.pdf"),
  plot    =fig2,
  width   =8,
  height  =5,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

## C. Figure 3

### i. Core taxa by site

```{r}
taxon_counts_site_18S=read.csv(file.path(paths$results_tables, "coreMOTUs_site_18S.csv"), header=T)
taxon_counts_site_COI=read.csv(file.path(paths$results_tables, "coreMOTUs_site_COI.csv"), header=T)
taxon_counts_site=rbind(taxon_counts_site_18S, taxon_counts_site_COI)

# Remove NA phyla
taxon_counts_site=taxon_counts_site %>% filter(!is.na(phylum))

# Define manual order (same for both markers)
desired_order=c("Hemichordata","Chordata", "Ochrophyta","Annelida","Mollusca","Rhodophyta","Cnidaria","Chlorophyta","Porifera", "NA")


# Apply factor levels
taxon_counts_site=taxon_counts_site %>%
  mutate(phylum=factor(phylum, levels=desired_order))

# Remove NA phyla
taxon_counts_site=taxon_counts_site %>%
  filter(!is.na(phylum))

fig3=ggplot(taxon_counts_site, aes(x=phylum, y=n, fill=Site)) +
  geom_bar(stat="identity", position="stack") +
  labs(x="", y="Total MOTUs") + 
  My_Theme +
  theme(
    plot.margin=unit(c(5, 20, 5, 5), "pt"),
    axis.text.y=element_text(size=10, hjust=1),
    legend.position="none"
  ) +
  scale_fill_manual(values=cols.site) +
  facet_wrap(~ Marker)+
   scale_y_sqrt() +coord_flip() 
fig3
```

###ii. Export

```{r}
ggsave(
  filename=file.path(paths$results_figs, "FIGURE_3.pdf"),
  plot    =fig3,
  width   =7,
  height  =4,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

##D. Figure 5

### i. Alpha diversity data

```{r}
div.18S=read.csv(
  file.path(paths$results_tables, "alpha_div_18S_iNEXT.csv"),
  header=TRUE
)
div.18S$MARKER="18S"
div.18S$prop=div.18S$Richness / max(div.18S$Richness, na.rm=TRUE)

div.COI=read.csv(
  file.path(paths$results_tables, "alpha_div_COI_iNEXT.csv"),
  header=TRUE
)
div.COI$MARKER="COI"
div.COI$prop=div.COI$Richness / max(div.COI$Richness, na.rm=TRUE)

# Combine
data_alphadiv=rbind(div.18S, div.COI)

# ===============================
# Data preparation
# ===============================
alpha=subset(data_alphadiv, STEP != 1)
alpha$STEP=as.factor(alpha$STEP)

alpha$DATE=as.character(alpha$DATE)

# Parse as DD/MM/YYYY
alpha$DATE=as.Date(alpha$DATE, format="%d/%m/%Y")

# Create MM/YY label
alpha$MONTH_YEAR=format(alpha$DATE, "%m/%y")

# Order chronologically
alpha$MONTH_YEAR=factor(
  alpha$MONTH_YEAR,
  levels=unique(alpha$MONTH_YEAR[
    order(as.yearmon(alpha$MONTH_YEAR, "%m/%y"))
  ])
)

# ===============================
# SST summary
# ===============================
alpha_avg_SST=alpha %>%
  group_by(MONTH_YEAR) %>%
  summarise(mean_SST=mean(SST, na.rm=TRUE), .groups="drop")

```

###ii. Panel A

```{r}
# ===============================
# Shading indices (based on MM/YY)
# ===============================
shading_start_index =which(levels(alpha$MONTH_YEAR) == "09/19")
shading_end_index   =which(levels(alpha$MONTH_YEAR) == "11/19")

shading_start_index2=which(levels(alpha$MONTH_YEAR) == "09/20")
shading_end_index2  =which(levels(alpha$MONTH_YEAR) == "11/20")

# ===============================
# Vertical line positions
# ===============================
step_levels=levels(alpha$MONTH_YEAR)
vline_positions=seq(1.5, length(step_levels) - 0.5, by=1)

# ===============================
# Plot
# ===============================
richnesstimestep=ggplot(
  alpha,
  aes(x=MONTH_YEAR, y=prop, fill=MARKER)
) +

  # Shaded periods
  annotate(
    "rect",
    xmin=shading_start_index - 0.5,
    xmax=shading_end_index + 0.5,
    ymin=-Inf, ymax=Inf,
    fill="grey55", alpha=0.3
  ) +
  annotate(
    "rect",
    xmin=shading_start_index2 - 0.5,
    xmax=shading_end_index2 + 0.5,
    ymin=-Inf, ymax=Inf,
    fill="grey55", alpha=0.3
  ) +

  # Vertical separators
  geom_vline(
    xintercept=vline_positions,
    linetype="dashed",
    color="grey"
  ) +

  # SST line (scaled to richness axis)
  geom_line(
    data=alpha_avg_SST,
    aes(
      x=MONTH_YEAR,
      y=(mean_SST - 18) / (28 - 18),
      group=1
    ),
    color="black",
    size=0.8, inherit.aes=F
  ) +

  # Distributions
  geom_violin(alpha=0.5, scale="width") +
  geom_boxplot(
    width=0.2,
    position=position_dodge(width=0.8),
    outlier.shape=NA,
    alpha=0.7
  ) +

  # Axes
  scale_y_continuous(
    name="Richness",
    sec.axis=sec_axis(
      ~ . * (28 - 18) + 18,
      name="SST (°C)",
      breaks=seq(18, 28, by=1)
    )
  ) +

  scale_fill_manual(values=cols.marker) +
  labs(x="", y="Richness") +

  theme(
    legend.position="none",
    axis.text.x=element_text(angle=45, hjust=1)
  )
```

###iii. Turnover data

```{r}
# Read the data
turnover.18S=read.csv(file.path(paths$results_tables, "turnover_18S.csv"), header=T)
turnover.COI=read.csv(file.path(paths$results_tables, "turnover_COI.csv"), header=T)

# Combine the data
turnover.site=rbind(turnover.18S, turnover.COI)

# Reorder SITE so "Overall" is the first level
turnover.site$SITE=fct_relevel(turnover.site$SITE, "Overall")

# Define colors for sites
site_levels=levels(turnover.site$SITE)
site_colors=c("Overall"="black", setNames(cols.site, site_levels[-1]))

# Assign final color based on Measure and SITE
turnover.site$final_color=ifelse(
  turnover.site$Measure == "Total",
  ifelse(turnover.site$SITE == "Overall", "Overall", as.character(turnover.site$SITE)),
  as.character(turnover.site$Measure)
)

# Convert MonthYear to Date using proper parsing
turnover.site$MonthYearDate=as.Date(paste0("01/", turnover.site$MonthYear), format="%d/%m/%y")

# Order data by MonthYearDate and update MonthYear factor levels
turnover.site=turnover.site[order(turnover.site$MonthYearDate), ]
turnover.site$MonthYear=factor(turnover.site$MonthYear, levels=unique(turnover.site$MonthYear))
```

### iv. Panel B

```{r}

# Select every other label for x-axis
labels_to_show=levels(turnover.site$MonthYear)[seq(1, length(levels(turnover.site$MonthYear)), by=2)]

# Identify the index positions for the shading range
shading_start_index=which(levels(turnover.site$MonthYear) == "09/19")
shading_end_index=which(levels(turnover.site$MonthYear) == "11/19")

shading_start_index2=which(levels(turnover.site$MonthYear) == "09/20")
shading_end_index2=which(levels(turnover.site$MonthYear) == "11/20")

labels_to_show=levels(turnover.site$MonthYear)[seq(1, length(levels(turnover.site$MonthYear)), by=3)]


# Create the plot with background shading
turnover.site.plot=ggplot(turnover.site, aes(x=MonthYear, y=value, color=final_color, linetype=Measure, group=interaction(Measure, SITE), shape=Measure)) +
  annotate("rect", xmin=shading_start_index-2, xmax=shading_end_index, ymin=-Inf, ymax=Inf, 
           fill="grey95", alpha=1) +
  annotate("rect", xmin=shading_start_index2, xmax=shading_end_index2, ymin=-Inf, ymax=Inf, 
           fill="grey95", alpha=1) + geom_vline(xintercept=c(shading_end_index2, shading_end_index, shading_start_index2), 
             linetype="dashed", color="grey80", size=0.7, alpha=0.6)+
  geom_line(aes(linewidth=Measure)) +  # Apply linewidth mapping
  ylab("Turnover")+
  xlab("") +
  theme_bw(base_size=18) +
  scale_x_discrete(
    breaks=labels_to_show,  
    labels=labels_to_show  
  ) +
  facet_grid(Marker ~ SITE) +
  My_Theme +
  scale_color_manual(values=c(cols.site, site_colors, "Appearance"="black", "Disappearance"="grey33")) +  # Change color of "Disappearance" to dark grey
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position="none") +
  scale_linetype_manual(values=c("Total"="solid", "Appearance"="solid", "Disappearance"="dotdash")) +
  scale_shape_manual(values=c("Total"=19, "Appearance"=16, "Disappearance"=15)) + 
  scale_linewidth_manual(values=c("Total"=1.5, "Appearance"=0.5, "Disappearance"=0.5)) +
  theme(legend.position="none")
```

### v. Export

```{r}
fig5=ggarrange(richnesstimestep, turnover.site.plot, ncol=1, labels="AUTO", font.label=list(size=18, color="black", face="bold", family="Arial"))
fig5
ggsave(
  filename=file.path(paths$results_figs, "FIGURE_5.pdf"),
  plot    =fig5,
  width   =9,
  height  =8,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

## E. Figure 4 GAM

###i. Prepare data

#### a. Richness

```{r}
alpha.mod=alpha
# Create prop2 scaled away from exact 0 and 1 - Smithson & Verkuilen correction
alpha.mod$prop2=(alpha.mod$prop * (nrow(alpha.mod) - 1) + 0.5) / nrow(alpha.mod)

# Step 1: Scale continuous predictors
alpha.mod$SST_scaled=scale(alpha.mod$SST)
alpha.mod$tide_scaled=scale(alpha.mod$tide_height_m)
alpha.mod$wave_scaled=scale(alpha.mod$WVHT)
alpha.mod$wind_scaled=scale(alpha.mod$WSPD)
alpha.mod$illum_scaled=scale(alpha.mod$illum)
alpha.mod$SITE=as.factor(alpha.mod$SITE)
alpha.mod$MARKER=as.factor(alpha.mod$MARKER)
alpha.mod$SEASON=as.factor(alpha.mod$SEASON)
```

####b. eDNA Index 

```{r}
data.long.18S=read.csv(file.path(paths$data_processed, "eDNA_data_long_transformed_18S.csv"), header=T)
data.long.18S$MARKER="18S"

data.long.COI=read.csv(file.path(paths$data_processed, "eDNA_data_long_transformed_COI.csv"), header=T)
data.long.COI$MARKER="COI"

data.long.all=rbind(data.long.18S, data.long.COI)

data.long.top5=subset(data.long.all, phylum %in% c("Arthropoda", "Annelida", "Cnidaria", "Mollusca", "Porifera")) #subsetting top phylum only, to reduce computation time of eDNA index GAMM

# Smithson & Verkuilen correction for beta family model fit
data.long.top5$value2=(data.long.top5$value * (nrow(data.long.top5) - 1) + 0.5) / nrow(data.long.top5)

data.long.top5=data.long.top5 %>%
  mutate(
    SST_scaled=scale(SST),
    tide_scaled=scale(tide_height_m),
    wave_scaled=scale(WVHT),
    wind_scaled=scale(WSPD),
    illum_scaled=scale(illum)
  )

data.long.top5=data.long.top5 %>%
  mutate(
    SITE=as.factor(SITE),
    MARKER=as.factor(MARKER),
    SEASON=as.factor(SEASON)
  )

phylum_counts=data.long.top5 %>%
  # Replace blank or NA phylum with "Unknown"
  mutate(phylum=ifelse(phylum == "" | is.na(phylum), "Unknown", phylum)) %>%
  mutate(phylum=factor(phylum))
```

### ii. Pooled GAMMs

#### a. Models

```{r}
knots=4 #knots for GAM smoothers

### Richness GAM ###########################################
model_beta_gam=mgcv::gam(prop2 ~ SEASON + MARKER +  s(SST_scaled, k=knots, bs="ts") + s(wind_scaled, k=knots, bs="ts") + s(wave_scaled, k=knots, bs="ts") +s(tide_scaled, k=knots, bs="ts") + s(degrees, k=knots, bs="cc") + s(MWD, k=knots, bs="cc") + s(WDIR, k=knots, bs="cc") + s(SITE, bs="re"),data=alpha.mod,family=betar(link="logit"),method="REML",select=TRUE)
saveRDS(model_beta_gam,file.path(paths$results_rds,"GAMMmodel_richness.rds"))
############################################################


### eDNA Index GAM #####################################################
model_beta_reads_gam=mgcv::gam(value2 ~ SEASON + MARKER + s(SST_scaled, k=knots, bs="ts") + s(wind_scaled, k=knots, bs="ts") + s(wave_scaled, k=knots, bs="ts") + s(tide_scaled, k=knots, bs="ts") + s(degrees, k=knots, bs="cc") + s(MWD, k=knots, bs="cc") + s(WDIR, k=knots, bs="cc") + s(SITE, bs="re") + s(phylum, bs="re"), data=phylum_counts, family=betar(link="logit"),method="REML",select=TRUE)
saveRDS(model_beta_reads_gam,file.path(paths$results_rds,"GAMMmodel_reads.rds"))
#######################################################################

##RICHNESS OUTPUT
model_beta_gam=readRDS(file=file.path(paths$results_rds,"GAMMmodel_richness.rds"))
gam_sum_richness=summary(model_beta_gam)

# Create parametric coefficients table with consistent column names
p_table_richness=data.frame(
  Term=rownames(gam_sum_richness$p.table),
  Estimate=round(gam_sum_richness$p.table[,1], 3),
  Std_Error=round(gam_sum_richness$p.table[,2], 3),
  Statistic=round(gam_sum_richness$p.table[,3], 2),
  P_value=gam_sum_richness$p.table[,4],
  Significance=ifelse(gam_sum_richness$p.table[,4] < 0.001, "***",
                 ifelse(gam_sum_richness$p.table[,4] < 0.01, "**",
                 ifelse(gam_sum_richness$p.table[,4] < 0.05, "*",
                 ifelse(gam_sum_richness$p.table[,4] < 0.1, ".", "")))),
  edf=NA,
  Ref_df=NA,
  Chi_sq=NA,
  stringsAsFactors=FALSE
)

# Clean up P_value column
p_table_richness$P_value=ifelse(p_table_richness$P_value < 0.001, "<0.001",
                            round(p_table_richness$P_value, 3))

# Create smooth terms table with the SAME column names
s_table_richness=data.frame(
  Term=rownames(gam_sum_richness$s.table),
  Estimate=round(gam_sum_richness$s.table[,1], 2),  # edf
  Std_Error=round(gam_sum_richness$s.table[,2], 2),  # Ref.df
  Statistic=round(gam_sum_richness$s.table[,3], 2),  # Chi.sq
  P_value=gam_sum_richness$s.table[,4],
  Significance=ifelse(gam_sum_richness$s.table[,4] < 0.001, "***",
                 ifelse(gam_sum_richness$s.table[,4] < 0.01, "**",
                 ifelse(gam_sum_richness$s.table[,4] < 0.05, "*",
                 ifelse(gam_sum_richness$s.table[,4] < 0.1, ".", "")))),
  edf=round(gam_sum_richness$s.table[,1], 2),
  Ref_df=round(gam_sum_richness$s.table[,2], 2),
  Chi_sq=round(gam_sum_richness$s.table[,3], 2),
  stringsAsFactors=FALSE
)

# Clean up P_value column for smooth terms
s_table_richness$P_value=ifelse(s_table_richness$P_value < 0.001, "<0.001",
                            round(s_table_richness$P_value, 3))

gam_summarytable_richness=rbind(
  data.frame(Type="Parametric", p_table_richness),
  data.frame(Type="Smooth", s_table_richness)
)

# Add model fit statistics
fit_stats_richness=data.frame(
  Type="Model Fit",
  Term=c("R-squared", "Deviance_explained", "REML_score", "n"),
  Estimate=c(round(gam_sum_richness$r.sq, 3),
               round(gam_sum_richness$dev.expl, 3),
               round(gam_sum_richness$sp.criterion, 1),
               gam_sum_richness$n),
  Std_Error=NA,
  Statistic=NA,
  P_value=NA,
  Significance=NA,
  edf=NA,
  Ref_df=NA,
  Chi_sq=NA,
  stringsAsFactors=FALSE
)

gam_summarytable_richness=rbind(gam_summarytable_richness, fit_stats_richness)

print(gam_summarytable_richness)

write.csv(gam_summarytable_richness, file=file.path(paths$results_tables,"TABLE_GAMM_richness_results.csv"), row.names=FALSE, na="")

##READS OUTPUT
model_beta_reads_gam=readRDS(file=file.path(paths$results_rds,"GAMMmodel_reads.rds"))
gam_sum_reads=summary(model_beta_reads_gam)
# Create parametric coefficients table with consistent column names
p_table=data.frame(
  Term=rownames(gam_sum_reads$p.table),
  Estimate=round(gam_sum_reads$p.table[,1], 3),
  Std_Error=round(gam_sum_reads$p.table[,2], 3),
  Statistic=round(gam_sum_reads$p.table[,3], 2),
  P_value=gam_sum_reads$p.table[,4],
  Significance=ifelse(gam_sum_reads$p.table[,4] < 0.001, "***",
                 ifelse(gam_sum_reads$p.table[,4] < 0.01, "**",
                 ifelse(gam_sum_reads$p.table[,4] < 0.05, "*",
                 ifelse(gam_sum_reads$p.table[,4] < 0.1, ".", "")))),
  edf=NA,
  Ref_df=NA,
  Chi_sq=NA,
  stringsAsFactors=FALSE
)

# Clean up P_value column
p_table$P_value=ifelse(p_table$P_value < 0.001, "<0.001",
                   round(p_table$P_value, 3))

# Create smooth terms table with the SAME column names
s_table=data.frame(
  Term=rownames(gam_sum_reads$s.table),
  Estimate=round(gam_sum_reads$s.table[,1], 2),  # edf
  Std_Error=round(gam_sum_reads$s.table[,2], 2),  # Ref.df
  Statistic=round(gam_sum_reads$s.table[,3], 2),  # Chi.sq
  P_value=gam_sum_reads$s.table[,4],
  Significance=ifelse(gam_sum_reads$s.table[,4] < 0.001, "***",
                 ifelse(gam_sum_reads$s.table[,4] < 0.01, "**",
                 ifelse(gam_sum_reads$s.table[,4] < 0.05, "*",
                 ifelse(gam_sum_reads$s.table[,4] < 0.1, ".", "")))),
  edf=round(gam_sum_reads$s.table[,1], 2),
  Ref_df=round(gam_sum_reads$s.table[,2], 2),
  Chi_sq=round(gam_sum_reads$s.table[,3], 2),
  stringsAsFactors=FALSE
)

# Clean up P_value column for smooth terms
s_table$P_value=ifelse(s_table$P_value < 0.001, "<0.001",
                   round(s_table$P_value, 3))

gam_summarytable_reads=rbind(
  data.frame(Type="Parametric", p_table),
  data.frame(Type="Smooth", s_table)
)

# Add model fit statistics
fit_stats=data.frame(
  Type="Model Fit",
  Term=c("R-squared", "Deviance_explained", "REML_score", "n"),
  Estimate=c(round(gam_sum_reads$r.sq, 3),
               round(gam_sum_reads$dev.expl, 3),
               round(gam_sum_reads$sp.criterion, 1),
               gam_sum_reads$n),
  Std_Error=NA,
  Statistic=NA,
  P_value=NA,
  Significance=NA,
  edf=NA,
  Ref_df=NA,
  Chi_sq=NA,
  stringsAsFactors=FALSE
)

gam_summarytable_reads=rbind(gam_summarytable_reads, fit_stats)

print(gam_summarytable_reads)

write.csv(gam_summarytable_reads, file=file.path(paths$results_tables, "TABLE_GAMM_reads_results.csv"), row.names=FALSE, na="")
```

#### b. Generate predictor fits

```{r}
predict_linear_gam=function(model, predictor, model_label,
                               ref_data, metadata=NULL,
                               n_points=500, color_line="black", 
                               linetype_line="solid") {
  
  # Realistic range for scaled predictor
  pred_seq=seq(min(ref_data[[predictor]], na.rm=TRUE),
                  max(ref_data[[predictor]], na.rm=TRUE),
                  length.out=n_points)
  
  # Get reference levels for factors
  if("SITE" %in% names(ref_data)) {
    common_site=names(sort(table(ref_data$SITE), decreasing=TRUE))[1]
  } else {
    common_site=NA
  }
  
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Template newdata with fixed reference values
  newdata=data.frame(
    SST_scaled=0,
    wind_scaled=0,
    wave_scaled=0,
    tide_scaled=0,
    degrees=180,
    MWD=180,
    WDIR=180,
    SEASON=factor("Dry", levels=levels(ref_data$SEASON)),
    SITE=factor(common_site, levels=levels(ref_data$SITE)),
    MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
  )
  
  if("phylum" %in% names(ref_data)) {
    newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
  }
  
  newdata=newdata[rep(1, n_points), ]
  newdata[[predictor]]=pred_seq
  rownames(newdata)=NULL
  
  # GAM predictions
  pred_link=tryCatch({
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  }, error=function(e) {
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  })
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  # Direct calculation
  newdata$fit=plogis(pred_link$fit)
  newdata$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata$model=model_label
  newdata$color_line=color_line
  newdata$linetype_line=linetype_line
  
  # Back-transform scaling to original units
  scaling_info=list(
    SST_scaled=list(orig="SST", mean=mean(ref_data$SST, na.rm=TRUE), sd=sd(ref_data$SST, na.rm=TRUE)),
    wind_scaled=list(orig="WSPD", mean=mean(ref_data$WSPD, na.rm=TRUE), sd=sd(ref_data$WSPD, na.rm=TRUE)),
    wave_scaled=list(orig="WVHT", mean=mean(ref_data$WVHT, na.rm=TRUE), sd=sd(ref_data$WVHT, na.rm=TRUE)),
    tide_scaled=list(orig="tide_height_m", mean=mean(ref_data$tide_height_m, na.rm=TRUE), sd=sd(ref_data$tide_height_m, na.rm=TRUE))
  )
  
  if(predictor %in% names(scaling_info)) {
    newdata[[paste0(predictor, "_orig")]]=
      newdata[[predictor]] * scaling_info[[predictor]]$sd + scaling_info[[predictor]]$mean
  } else {
    newdata[[paste0(predictor, "_orig")]]=newdata[[predictor]]
  }
  
  # Observed data for histogram
  obs_data=NULL
  if(!is.null(metadata) & predictor %in% names(scaling_info)) {
    obs_var=scaling_info[[predictor]]$orig
    obs_data=metadata %>%
      dplyr::select(dplyr::all_of(obs_var)) %>%
      dplyr::rename(obs_value=dplyr::all_of(obs_var)) %>%
      dplyr::filter(!is.na(obs_value))
  }
  
  return(list(predictions=newdata, observed=obs_data))
}

predict_cyclic_gam=function(model, predictor, model_label,
                               ref_data, metadata=NULL, obs_var=NULL,
                               n_points=500, color_line="black", 
                               linetype_line="solid", polar=FALSE) {
  
  # Sequence for cyclic predictor (0 to 360 degrees)
  angle_seq=seq(0, 360, length.out=n_points)
  
  # Get reference levels for factors
  if("SITE" %in% names(ref_data)) {
    common_site=names(sort(table(ref_data$SITE), decreasing=TRUE))[1]
  } else {
    common_site=NA
  }
  
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Template newdata
  newdata=data.frame(
    SST_scaled=0,
    wind_scaled=0,
    wave_scaled=0,
    tide_scaled=0,
    degrees=180,
    MWD=180,
    WDIR=180,
    SEASON=factor("Dry", levels=levels(ref_data$SEASON)),
    SITE=factor(common_site, levels=levels(ref_data$SITE)),
    MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
  )
  
  if("phylum" %in% names(ref_data)) {
    newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
  }
  
  newdata=newdata[rep(1, n_points), ]
  rownames(newdata)=NULL
  
  # Vary chosen cyclic predictor
  if(predictor == "WDIR") {
    newdata$WDIR=angle_seq
  } else if(predictor == "MWD") {
    newdata$MWD=angle_seq
  } else if(predictor == "lunar") {
    newdata$degrees=angle_seq
  }
  
  newdata$direction_deg=angle_seq
  
  # GAM predictions
  pred_link=tryCatch({
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  }, error=function(e) {
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  })
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  # Direct calculation
  newdata$fit=plogis(pred_link$fit)
  newdata$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata$model=model_label
  newdata$color_line=color_line
  newdata$linetype_line=linetype_line
  
  # Observed frequency overlay
  obs_plot=NULL
  if(!is.null(metadata) & !is.null(obs_var)) {
    obs_plot=metadata %>%
      dplyr::select(dplyr::all_of(obs_var)) %>%
      dplyr::filter(!is.na(.data[[obs_var]])) %>%
      dplyr::mutate(bin=cut(.data[[obs_var]], breaks=seq(0, 360, length.out=51), include.lowest=TRUE)) %>%
      dplyr::group_by(bin) %>%
      dplyr::summarise(freq=dplyr::n()) %>%
      dplyr::mutate(direction_deg=seq(0, 360, length.out=dplyr::n()+1)[-1],
                    freq_scaled=freq / max(freq))
  }
  
  return(list(predictions=newdata, obs=obs_plot))
}

predict_categorical_gam=function(model, predictor, model_label,
                                    ref_data, color_line="black", 
                                    linetype_line="solid") {
  
  # Get factor levels
  pred_levels=levels(ref_data[[predictor]])
  
  # Get reference levels for other factors
  if("SITE" %in% names(ref_data)) {
    common_site=names(sort(table(ref_data$SITE), decreasing=TRUE))[1]
  } else {
    common_site=NA
  }
  
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Template newdata
  newdata=data.frame(
    SST_scaled=0,
    wind_scaled=0,
    wave_scaled=0,
    tide_scaled=0,
    degrees=180,
    MWD=180,
    WDIR=180,
    SEASON=factor(pred_levels[1], levels=pred_levels),
    SITE=factor(common_site, levels=levels(ref_data$SITE)),
    MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
  )
  
  if("phylum" %in% names(ref_data)) {
    newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
  }
  
  newdata=newdata[rep(1, length(pred_levels)), ]
  newdata[[predictor]]=factor(pred_levels, levels=pred_levels)
  rownames(newdata)=NULL
  
  # GAM predictions
  pred_link=tryCatch({
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  }, error=function(e) {
    predict(model, newdata=newdata, type="link", se.fit=TRUE)
  })
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  # Direct calculation
  newdata$fit=plogis(pred_link$fit)
  newdata$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata$model=model_label
  newdata$color_line=color_line
  newdata$linetype_line=linetype_line
  
  # Add numeric version for plotting
  newdata$SEASON_orig=as.numeric(newdata$SEASON)
  
  return(newdata)
}

# Wind direction
pred_wdir_rich_gam=predict_cyclic_gam(
  model=model_beta_gam,
  predictor="WDIR",
  model_label="Detected richness",
  ref_data=alpha.mod,
  metadata=alpha.mod,
  obs_var="WDIR",
  color_line="#1985a1",
  linetype_line="dashed"
)

pred_wdir_edna_gam=predict_cyclic_gam(
  model=model_beta_reads_gam,
  predictor="WDIR",
  model_label="eDNA index",
  ref_data=phylum_counts,
  metadata=phylum_counts,
  obs_var="WDIR",
  color_line="#1985a1",
  linetype_line="solid"
)

# Wave direction
pred_mwd_rich_gam=predict_cyclic_gam(
  model=model_beta_gam,
  predictor="MWD",
  model_label="Detected richness",
  ref_data=alpha.mod,
  metadata=alpha.mod,
  obs_var="MWD",
  color_line="black",
  linetype_line="dashed"
)

pred_mwd_edna_gam=predict_cyclic_gam(
  model=model_beta_reads_gam,
  predictor="MWD",
  model_label="eDNA index",
  ref_data=phylum_counts,
  metadata=phylum_counts,
  obs_var="MWD",
  color_line="#1985a1",
  linetype_line="solid"
)

# Lunar angle
pred_lunar_rich_gam=predict_cyclic_gam(
  model=model_beta_gam,
  predictor="lunar",
  model_label="Detected richness",
  ref_data=alpha.mod,
  metadata=alpha.mod,
  obs_var="degrees",
  color_line="#1985a1",
  linetype_line="dashed"
)

pred_lunar_edna_gam=predict_cyclic_gam(
  model=model_beta_reads_gam,
  predictor="lunar",
  model_label="eDNA index",
  ref_data=phylum_counts,
  metadata=phylum_counts,
  obs_var="degrees",
  color_line="black",
  linetype_line="solid"
)

# Scaled predictors
scaled_predictors=c("SST_scaled", "wind_scaled", "wave_scaled", "tide_scaled")

pred_scaled_rich_list=list()
pred_scaled_edna_list=list()

for(pred in scaled_predictors) {
    pred_scaled_rich_list[[pred]]=predict_linear_gam(
    model=model_beta_gam,
    predictor=pred,
    model_label="Detected richness",
    ref_data=alpha.mod,
    metadata=alpha.mod,
    color_line="#1985a1",
    linetype_line="dashed"
  )
  
  pred_scaled_edna_list[[pred]]=predict_linear_gam(
    model=model_beta_reads_gam,
    predictor=pred,
    model_label="eDNA index",
    ref_data=phylum_counts,
    metadata=phylum_counts,
    color_line="black",
    linetype_line="solid"
  )
}

# Season predictions
pred_season_rich_gam=predict_categorical_gam(
  model=model_beta_gam,
  predictor="SEASON",
  model_label="Detected richness",
  ref_data=alpha.mod,
  color_line="#1985a1",
  linetype_line="dashed"
)

pred_season_edna_gam=predict_categorical_gam(
  model=model_beta_reads_gam,
  predictor="SEASON",
  model_label="eDNA index",
  ref_data=phylum_counts,
  color_line="black",
  linetype_line="solid"
)
```

#### c. Plot Figure 4

```{r}
# For Richness: find global min and max across ALL predictors
all_richness_fits=c()

# Collect all Richness predictions from all predictors
for(pred in scaled_predictors) {
  all_richness_fits=c(all_richness_fits, 
                         pred_scaled_rich_list[[pred]]$predictions$fit,
                         pred_scaled_rich_list[[pred]]$predictions$lower,
                         pred_scaled_rich_list[[pred]]$predictions$upper)
}

# Add cyclic predictors
all_richness_fits=c(all_richness_fits,
                       pred_wdir_rich_gam$predictions$fit,
                       pred_wdir_rich_gam$predictions$lower,
                       pred_wdir_rich_gam$predictions$upper,
                       pred_mwd_rich_gam$predictions$fit,
                       pred_mwd_rich_gam$predictions$lower,
                       pred_mwd_rich_gam$predictions$upper,
                       pred_lunar_rich_gam$predictions$fit,
                       pred_lunar_rich_gam$predictions$lower,
                       pred_lunar_rich_gam$predictions$upper)

# Add season
all_richness_fits=c(all_richness_fits,
                       pred_season_rich_gam$fit,
                       pred_season_rich_gam$lower,
                       pred_season_rich_gam$upper)

# Calculate global Richness range
global_richness_range=range(all_richness_fits, na.rm=TRUE)
global_richness_padding=diff(global_richness_range) * 0.15
global_richness_ylim=c(max(0, global_richness_range[1] - global_richness_padding), 
                          min(1, global_richness_range[2] + global_richness_padding))

# For eDNA Index: find global min and max across ALL predictors
all_edna_fits=c()

# Collect all eDNA predictions from all predictors
for(pred in scaled_predictors) {
  all_edna_fits=c(all_edna_fits, 
                     pred_scaled_edna_list[[pred]]$predictions$fit,
                     pred_scaled_edna_list[[pred]]$predictions$lower,
                     pred_scaled_edna_list[[pred]]$predictions$upper)
}

# Add cyclic predictors
all_edna_fits=c(all_edna_fits,
                   pred_wdir_edna_gam$predictions$fit,
                   pred_wdir_edna_gam$predictions$lower,
                   pred_wdir_edna_gam$predictions$upper,
                   pred_mwd_edna_gam$predictions$fit,
                   pred_mwd_edna_gam$predictions$lower,
                   pred_mwd_edna_gam$predictions$upper,
                   pred_lunar_edna_gam$predictions$fit,
                   pred_lunar_edna_gam$predictions$lower,
                   pred_lunar_edna_gam$predictions$upper)

# Add season
all_edna_fits=c(all_edna_fits,
                   pred_season_edna_gam$fit,
                   pred_season_edna_gam$lower,
                   pred_season_edna_gam$upper)

# Calculate global eDNA range
global_edna_range=range(all_edna_fits, na.rm=TRUE)
global_edna_padding=diff(global_edna_range) * 0.15
global_edna_ylim=c(max(0, global_edna_range[1] - global_edna_padding), 
                      min(1, global_edna_range[2] + global_edna_padding))

## PLOTTING FUNCTIONS

create_cyclic_combined_plot=function(rich_pred, edna_pred, rich_obs, edna_obs, 
                                        predictor_name, x_label,
                                        global_richness_ylim, global_edna_ylim) {
  
  # Prepare data
  rich_data=rich_pred
  edna_data=edna_pred
  
  rich_data$Response="Richness"
  edna_data$Response="eDNA Index"
  
  # Scale eDNA data to Richness scale for plotting
  edna_data$fit_scaled=scales::rescale(edna_data$fit, from=global_edna_ylim, to=global_richness_ylim)
  edna_data$lower_scaled=scales::rescale(edna_data$lower, from=global_edna_ylim, to=global_richness_ylim)
  edna_data$upper_scaled=scales::rescale(edna_data$upper, from=global_edna_ylim, to=global_richness_ylim)
  
  # Get observation data for histogram
  obs_data=rich_obs$obs
  if(is.null(obs_data) && !is.null(edna_obs$obs)) {
    obs_data=edna_obs$obs
  }
  
  # Create histogram data
  if(!is.null(obs_data)) {
    raw_angles=rep(obs_data$direction_deg, times=round(obs_data$freq_scaled * 100))
    hist_obj=hist(raw_angles, breaks=seq(0, 360, length.out=36), plot=FALSE)
    hist_df=data.frame(
      direction_deg=hist_obj$mids,
      density=hist_obj$density,
      norm_density=hist_obj$density / max(hist_obj$density)
    )
  } else {
    hist_df=data.frame(direction_deg=numeric(0), norm_density=numeric(0))
  }
  
  # Transformation function for secondary axis
  trans_formula=function(x) {
    scales::rescale(x, from=global_richness_ylim, to=global_edna_ylim)
  }
  
  # Create plot
  plot=ggplot() +
    # Histogram
    {if(nrow(hist_df) > 0) geom_bar(data=hist_df, 
       aes(x=direction_deg, y=norm_density * (global_richness_ylim[2] - global_richness_ylim[1]) * 0.15 + global_richness_ylim[1]), 
       stat="identity", fill="tan", alpha=0.25, width=10)} +
    # Richness
    geom_ribbon(data=rich_data,
                aes(x=direction_deg, ymin=lower, ymax=upper),
                fill="#1985a1", alpha=0.3) +
    geom_line(data=rich_data,
              aes(x=direction_deg, y=fit, color="Richness", linetype="Richness"),
              size=1.2) +
    # eDNA (rescaled)
    geom_ribbon(data=edna_data,
                aes(x=direction_deg, ymin=lower_scaled, ymax=upper_scaled),
                fill="black", alpha=0.3) +
    geom_line(data=edna_data,
              aes(x=direction_deg, y=fit_scaled, color="eDNA Index", linetype="eDNA Index"),
              size=1.2) +
    scale_y_continuous(
      name="Predicted Richness",
      limits=global_richness_ylim,
      labels=scales::number_format(accuracy=0.01),
      sec.axis=sec_axis(
        ~ trans_formula(.),
        name="Predicted eDNA Index",
        labels=scales::number_format(accuracy=0.001)
      )
    ) +
    scale_color_manual(values=c("Richness"="#1985a1", "eDNA Index"="black")) +
    scale_linetype_manual(values=c("Richness"="dashed", "eDNA Index"="solid")) +
    labs(x=x_label, 
         y=NULL,
         color="Response", 
         linetype="Response") +
    My_Theme +
    theme(
      legend.position="bottom",
      axis.title.y.left=element_text(color="#1985a1", face="bold"),
      axis.text.y.left=element_text(color="#1985a1"),
      axis.title.y.right=element_text(color="black", face="bold"),
      axis.text.y.right=element_text(color="black")
    )
  
  return(plot)
}

create_scaled_combined_plot=function(rich_pred_list, edna_pred_list, 
                                        rich_obs_list, edna_obs_list,
                                        predictor_name, x_label, predictor_key,
                                        global_richness_ylim, global_edna_ylim) {
  
  # Extract predictions
  rich_pred=rich_pred_list[[predictor_key]]$predictions
  edna_pred=edna_pred_list[[predictor_key]]$predictions
  
  # Get observed data
  rich_obs=rich_obs_list[[predictor_key]]$observed
  edna_obs=edna_obs_list[[predictor_key]]$observed
  
  # Get the original scale column name
  orig_col=paste0(predictor_key, "_orig")
  
  # Scale eDNA data to Richness scale
  edna_pred$fit_scaled=scales::rescale(edna_pred$fit, from=global_edna_ylim, to=global_richness_ylim)
  edna_pred$lower_scaled=scales::rescale(edna_pred$lower, from=global_edna_ylim, to=global_richness_ylim)
  edna_pred$upper_scaled=scales::rescale(edna_pred$upper, from=global_edna_ylim, to=global_richness_ylim)
  
  # Get observation data for histogram
  obs_data=rich_obs
  if(is.null(obs_data)) obs_data=edna_obs
  
  # Create normalized histogram
  if(!is.null(obs_data) && nrow(obs_data) > 0) {
    hist_obj=hist(obs_data$obs_value, breaks=30, plot=FALSE)
    hist_df=data.frame(
      x_value=hist_obj$mids,
      density=hist_obj$density,
      norm_density=hist_obj$density / max(hist_obj$density)
    )
  } else {
    hist_df=data.frame(x_value=numeric(0), norm_density=numeric(0))
  }
  
  # Transformation function for secondary axis
  trans_formula=function(x) {
    scales::rescale(x, from=global_richness_ylim, to=global_edna_ylim)
  }
  
  # Create plot
  plot=ggplot() +
    # Histogram
    {if(nrow(hist_df) > 0) geom_bar(data=hist_df, 
       aes(x=x_value, y=norm_density * (global_richness_ylim[2] - global_richness_ylim[1]) * 0.15 + global_richness_ylim[1]), 
       stat="identity", fill="tan", alpha=0.25, width=diff(hist_df$x_value)[1] * 0.9)} +
    # Richness
    geom_ribbon(data=rich_pred,
                aes(x=.data[[orig_col]], ymin=lower, ymax=upper),
                fill="#1985a1", alpha=0.3) +
    geom_line(data=rich_pred,
              aes(x=.data[[orig_col]], y=fit, color="Richness", linetype="Richness"),
              size=1.2) +
    # eDNA (rescaled)
    geom_ribbon(data=edna_pred,
                aes(x=.data[[orig_col]], ymin=lower_scaled, ymax=upper_scaled),
                fill="black", alpha=0.3) +
    geom_line(data=edna_pred,
              aes(x=.data[[orig_col]], y=fit_scaled, color="eDNA Index", linetype="eDNA Index"),
              size=1.2) +
    scale_y_continuous(
      name="Predicted Richness",
      limits=global_richness_ylim,
      labels=scales::number_format(accuracy=0.01),
      sec.axis=sec_axis(
        ~ trans_formula(.),
        name="Predicted eDNA Index",
        labels=scales::number_format(accuracy=0.001)
      )
    ) +
    scale_color_manual(values=c("Richness"="#1985a1", "eDNA Index"="black")) +
    scale_linetype_manual(values=c("Richness"="dashed", "eDNA Index"="solid")) +
    labs(x=x_label, 
         y=NULL,
         color="Response", 
         linetype="Response") +
    My_Theme +
    theme(
      legend.position="bottom",
      axis.title.y.left=element_text(color="#1985a1", face="bold"),
      axis.text.y.left=element_text(color="#1985a1"),
      axis.title.y.right=element_text(color="black", face="bold"),
      axis.text.y.right=element_text(color="black")
    )
  
  return(plot)
}


# Cyclic predictors
cyclic_wdir_plot=create_cyclic_combined_plot(
  pred_wdir_rich_gam$predictions, 
  pred_wdir_edna_gam$predictions,
  pred_wdir_rich_gam, 
  pred_wdir_edna_gam,
  "Wind Direction", "Wind direction (°)",
  global_richness_ylim, global_edna_ylim
)

cyclic_mwd_plot=create_cyclic_combined_plot(
  pred_mwd_rich_gam$predictions, 
  pred_mwd_edna_gam$predictions,
  pred_mwd_rich_gam, 
  pred_mwd_edna_gam,
  "Wave Direction", "Wave direction (°)",
  global_richness_ylim, global_edna_ylim
)

cyclic_lunar_plot=create_cyclic_combined_plot(
  pred_lunar_rich_gam$predictions, 
  pred_lunar_edna_gam$predictions,
  pred_lunar_rich_gam, 
  pred_lunar_edna_gam,
  "Lunar Angle", "Lunar angle (°)",
  global_richness_ylim, global_edna_ylim
)

# Scaled predictors
scaled_sst=create_scaled_combined_plot(
  pred_scaled_rich_list, pred_scaled_edna_list,
  pred_scaled_rich_list, pred_scaled_edna_list,
  "Sea Surface Temperature", "SST (°C)", "SST_scaled",
  global_richness_ylim, global_edna_ylim
)

scaled_wind=create_scaled_combined_plot(
  pred_scaled_rich_list, pred_scaled_edna_list,
  pred_scaled_rich_list, pred_scaled_edna_list,
  "Wind Speed", "Wind speed (m/s)", "wind_scaled",
  global_richness_ylim, global_edna_ylim
)

scaled_wave=create_scaled_combined_plot(
  pred_scaled_rich_list, pred_scaled_edna_list,
  pred_scaled_rich_list, pred_scaled_edna_list,
  "Wave Height", "Wave height (m)", "wave_scaled",
  global_richness_ylim, global_edna_ylim
)

scaled_tide=create_scaled_combined_plot(
  pred_scaled_rich_list, pred_scaled_edna_list,
  pred_scaled_rich_list, pred_scaled_edna_list,
  "Tide Height", "Tide height (m)", "tide_scaled",
  global_richness_ylim, global_edna_ylim
)

# Season plot
season_rich=pred_season_rich_gam
season_edna=pred_season_edna_gam

# Scale eDNA to richness scale
season_edna$fit_scaled=scales::rescale(season_edna$fit, from=global_edna_ylim, to=global_richness_ylim)
season_edna$lower_scaled=scales::rescale(season_edna$lower, from=global_edna_ylim, to=global_richness_ylim)
season_edna$upper_scaled=scales::rescale(season_edna$upper, from=global_edna_ylim, to=global_richness_ylim)

season_rich$Response="Richness"
season_edna$Response="eDNA Index"

# Transformation function for secondary axis
trans_formula=function(x) {
  scales::rescale(x, from=global_richness_ylim, to=global_edna_ylim)
}

# Create season plot with dual y-axes
plot_season_combined=ggplot() +
  # Richness points and error bars
  geom_point(data=season_rich,
             aes(x=SEASON, y=fit, shape=Response, color=Response),
             position=position_dodge(width=0.5), size=3) +
  geom_errorbar(data=season_rich,
                aes(x=SEASON, ymin=lower, ymax=upper, linetype=Response),
                position=position_dodge(width=0.5), width=0.2, size=0.8) +
  # eDNA points and error bars (scaled to richness range)
  geom_point(data=season_edna,
             aes(x=SEASON, y=fit_scaled, shape=Response, color=Response),
             position=position_dodge(width=0.5), size=3) +
  geom_errorbar(data=season_edna,
                aes(x=SEASON, ymin=lower_scaled, ymax=upper_scaled, linetype=Response),
                position=position_dodge(width=0.5), width=0.2, size=0.8) +
  scale_y_continuous(
    name="Predicted Richness",
    limits=global_richness_ylim,
    labels=scales::number_format(accuracy=0.01),
    sec.axis=sec_axis(
      ~ trans_formula(.),
      name="Predicted eDNA Index",
      labels=scales::number_format(accuracy=0.001)
    )
  ) +
  scale_color_manual(values=c("Richness"="#1985a1", "eDNA Index"="black")) +
  scale_linetype_manual(values=c("Richness"="dashed", "eDNA Index"="solid")) +
  scale_shape_manual(values=c("Richness"=16, "eDNA Index"=17)) +
  labs(x="Season", 
       y=NULL) +
  My_Theme +
  theme(
    legend.position="bottom",
    axis.title.y.left=element_text(color="#1985a1", face="bold"),
    axis.text.y.left=element_text(color="#1985a1"),
    axis.title.y.right=element_text(color="black", face="bold"),
    axis.text.y.right=element_text(color="black")
  )


GAM_scaled=ggarrange(
  scaled_wind, cyclic_wdir_plot, 
  scaled_wave, cyclic_mwd_plot, 
  scaled_tide, cyclic_lunar_plot, 
  scaled_sst, plot_season_combined,
  ncol=4, nrow=2,
  common.legend=TRUE,
  legend="bottom",
  labels="auto"
)

print(GAM_scaled)

ggsave(filename=file.path(paths$results_figs, "FIGURE_4_GAM.pdf"), plot    =GAM_scaled, width= 11, height=8, units="in", device=cairo_pdf, family="Arial")
```

### iii. Site-wise GAMMs

#### a. Models

```{r}
model_beta_gam_site=mgcv::gam(
  prop2 ~ SEASON + MARKER + 
    s(SST_scaled, SITE, bs="fs", k=knots) +
    s(wind_scaled, SITE, bs="fs", k=knots) +
    s(wave_scaled, SITE, bs="fs", k=knots) +
    s(tide_scaled, SITE, bs="fs", k=knots) +
    s(degrees, SITE, bs="fs", k=knots, xt=list(bs="cc")) +
    s(MWD, SITE, bs="fs", k=knots, xt=list(bs="cc")) +
    s(WDIR, SITE, bs="fs", k=knots, xt=list(bs="cc")),
  data=alpha.mod,
  family=betar(link="logit"),
  method="REML",
  select=TRUE
)
saveRDS(model_beta_gam_site,file.path(paths$results_rds,"GAMMmodel_richness_site.rds"))
model_beta_gam_site=readRDS(file.path(paths$results_rds,"GAMMmodel_richness_site.rds"))
summary(model_beta_gam_site)


### eDNA Index GAM
model_beta_reads_gam_site=mgcv::gam(
  value2 ~ SEASON + MARKER +
    s(SST_scaled, SITE, bs="fs", k=knots) +
    s(wind_scaled, SITE, bs="fs", k=knots) +
    s(wave_scaled, SITE, bs="fs", k=knots) +
    s(tide_scaled, SITE, bs="fs", k=knots) +
    s(degrees, SITE, bs="fs", k=knots, xt=list(bs="cc")) +
    s(MWD, SITE, bs="fs", k=knots, xt=list(bs="cc")) +
    s(WDIR, SITE, bs="fs", k=knots, xt=list(bs="cc")) +
    s(phylum, bs="re"),
  data=phylum_counts,
  family=betar(link="logit"),
  method="REML",
  select=TRUE
)
saveRDS(model_beta_reads_gam_site,file.path(paths$results_rds,"GAMMmodel_reads_site.rds"))
model_beta_reads_gam_site=readRDS(file.path(paths$results_rds,"GAMMmodel_reads_site.rds"))
summary(model_beta_reads_gam_site)
```

#### b. Generate predictions

```{r}
# Function for linear predictors (SST, wind, wave, tide)
predict_linear_gam_sitewise=function(model, predictor, model_label,
                               ref_data, metadata=NULL,
                               n_points=500) {
  
  # Realistic range for scaled predictor
  pred_seq=seq(min(ref_data[[predictor]], na.rm=TRUE),
                  max(ref_data[[predictor]], na.rm=TRUE),
                  length.out=n_points)
  
  # Get reference levels for factors
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Get all unique sites
  all_sites=unique(ref_data$SITE)
  
  # Create prediction data for all sites
  newdata_list=list()
  
  for(site in all_sites) {
    # Template newdata for this site
    newdata=data.frame(
      SST_scaled=0,
      wind_scaled=0,
      wave_scaled=0,
      tide_scaled=0,
      degrees=180,
      MWD=180,
      WDIR=180,
      SEASON=factor("Dry", levels=levels(ref_data$SEASON)),
      SITE=factor(site, levels=levels(ref_data$SITE)),
      MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
    )
    
    if("phylum" %in% names(ref_data)) {
      newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
    }
    
    newdata=newdata[rep(1, n_points), ]
    newdata[[predictor]]=pred_seq
    rownames(newdata)=NULL
    newdata_list[[site]]=newdata
  }
  
  # Combine all sites
  newdata_all=bind_rows(newdata_list, .id="SITE_site")
  
  # Predict for all sites 
  pred_link=predict(model, newdata=newdata_all, type="link", se.fit=TRUE)
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  # Calculate predictions
  newdata_all$fit=plogis(pred_link$fit)
  newdata_all$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata_all$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata_all$model=model_label
  
  # Back-transform scaling to original units
  scaling_info=list(
    SST_scaled=list(orig="SST", mean=mean(ref_data$SST, na.rm=TRUE), sd=sd(ref_data$SST, na.rm=TRUE)),
    wind_scaled=list(orig="WSPD", mean=mean(ref_data$WSPD, na.rm=TRUE), sd=sd(ref_data$WSPD, na.rm=TRUE)),
    wave_scaled=list(orig="WVHT", mean=mean(ref_data$WVHT, na.rm=TRUE), sd=sd(ref_data$WVHT, na.rm=TRUE)),
    tide_scaled=list(orig="tide_height_m", mean=mean(ref_data$tide_height_m, na.rm=TRUE), sd=sd(ref_data$tide_height_m, na.rm=TRUE))
  )
  
  if(predictor %in% names(scaling_info)) {
    newdata_all[[paste0(predictor, "_orig")]]=
      newdata_all[[predictor]] * scaling_info[[predictor]]$sd + scaling_info[[predictor]]$mean
  } else {
    newdata_all[[paste0(predictor, "_orig")]]=newdata_all[[predictor]]
  }
  
  return(newdata_all)
}

# Function for cyclic predictors (WDIR, MWD, lunar)
predict_cyclic_gam_sitewise=function(model, predictor, model_label,
                               ref_data, metadata=NULL, obs_var=NULL,
                               n_points=500, polar=FALSE) {
  
  # Sequence for cyclic predictor (0 to 360 degrees)
  angle_seq=seq(0, 360, length.out=n_points)
  
  # Get reference levels for factors
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Get all unique sites
  all_sites=unique(ref_data$SITE)
  
  # Create prediction data for all sites
  newdata_list=list()
  
  for(site in all_sites) {
    # Template newdata
    newdata=data.frame(
      SST_scaled=0,
      wind_scaled=0,
      wave_scaled=0,
      tide_scaled=0,
      degrees=180,
      MWD=180,
      WDIR=180,
      SEASON=factor("Dry", levels=levels(ref_data$SEASON)),
      SITE=factor(site, levels=levels(ref_data$SITE)),
      MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
    )
    
    if("phylum" %in% names(ref_data)) {
      newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
    }
    
    newdata=newdata[rep(1, n_points), ]
    rownames(newdata)=NULL
    
    # Vary chosen cyclic predictor
    if(predictor == "WDIR") {
      newdata$WDIR=angle_seq
    } else if(predictor == "MWD") {
      newdata$MWD=angle_seq
    } else if(predictor == "lunar") {
      newdata$degrees=angle_seq
    }
    
    newdata$direction_deg=angle_seq
    newdata_list[[site]]=newdata
  }
  
  # Combine all sites
  newdata_all=bind_rows(newdata_list, .id="SITE_site")
  
  # Predict for all sites
  pred_link=predict(model, newdata=newdata_all, type="link", se.fit=TRUE)
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  # Calculate predictions
  newdata_all$fit=plogis(pred_link$fit)
  newdata_all$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata_all$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata_all$model=model_label
  
  return(newdata_all)
}

# Function for categorical predictor (SEASON)
predict_categorical_gam_sitewise=function(model, predictor, model_label,
                                    ref_data) {
  
  # Get factor levels
  pred_levels=levels(ref_data[[predictor]])
  
  # Get reference levels for other factors
  if("MARKER" %in% names(ref_data)) {
    common_marker=names(sort(table(ref_data$MARKER), decreasing=TRUE))[1]
  } else {
    common_marker=NA
  }
  
  if("phylum" %in% names(ref_data)) {
    common_phylum=names(sort(table(ref_data$phylum), decreasing=TRUE))[1]
  } else {
    common_phylum=NA
  }
  
  # Get all unique sites
  all_sites=unique(ref_data$SITE)
  
  # Create prediction data for all sites and seasons
  newdata_list=list()
  
  for(site in all_sites) {
    for(season in pred_levels) {
      newdata=data.frame(
        SST_scaled=0,
        wind_scaled=0,
        wave_scaled=0,
        tide_scaled=0,
        degrees=180,
        MWD=180,
        WDIR=180,
        SEASON=factor(season, levels=pred_levels),
        SITE=factor(site, levels=levels(ref_data$SITE)),
        MARKER=factor(common_marker, levels=levels(ref_data$MARKER))
      )
      
      if("phylum" %in% names(ref_data)) {
        newdata$phylum=factor(common_phylum, levels=levels(ref_data$phylum))
      }
      
      newdata_list[[paste(site, season)]]=newdata
    }
  }
  
  # Combine all
  newdata_all=bind_rows(newdata_list, .id="id")
  newdata_all$SITE_site=gsub("_.*", "", newdata_all$id)
  newdata_all$SEASON_val=gsub(".*_", "", newdata_all$id)
  
  # Predict
  pred_link=predict(model, newdata=newdata_all, type="link", se.fit=TRUE)
  
  # Handle NA se.fit
  if(any(is.na(pred_link$se.fit))) {
    valid_se=pred_link$se.fit[!is.na(pred_link$se.fit)]
    if(length(valid_se) > 0) {
      mean_se=mean(valid_se)
    } else {
      mean_se=0.1
    }
    pred_link$se.fit[is.na(pred_link$se.fit)]=mean_se
  }
  
  newdata_all$fit=plogis(pred_link$fit)
  newdata_all$lower=plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata_all$upper=plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata_all$model=model_label
  
  return(newdata_all)
}

# Wind direction
pred_wdir_rich_gam=predict_cyclic_gam_sitewise(
  model=model_beta_gam,
  predictor="WDIR",
  model_label="Detected richness",
  ref_data=alpha.mod
)

pred_wdir_edna_gam=predict_cyclic_gam_sitewise(
  model=model_beta_reads_gam,
  predictor="WDIR",
  model_label="eDNA index",
  ref_data=phylum_counts
)

# Wave direction
pred_mwd_rich_gam=predict_cyclic_gam_sitewise(
  model=model_beta_gam,
  predictor="MWD",
  model_label="Detected richness",
  ref_data=alpha.mod
)

pred_mwd_edna_gam=predict_cyclic_gam_sitewise(
  model=model_beta_reads_gam,
  predictor="MWD",
  model_label="eDNA index",
  ref_data=phylum_counts
)

# Lunar angle
pred_lunar_rich_gam=predict_cyclic_gam_sitewise(
  model=model_beta_gam,
  predictor="lunar",
  model_label="Detected richness",
  ref_data=alpha.mod
)

pred_lunar_edna_gam=predict_cyclic_gam_sitewise(
  model=model_beta_reads_gam,
  predictor="lunar",
  model_label="eDNA index",
  ref_data=phylum_counts
)

# Scaled predictors
scaled_predictors=c("SST_scaled", "wind_scaled", "wave_scaled", "tide_scaled")

pred_scaled_rich_list=list()
pred_scaled_edna_list=list()

for(pred in scaled_predictors) {
  cat(sprintf("Generating predictions for %s...\n", pred))
  
  pred_scaled_rich_list[[pred]]=predict_linear_gam_sitewise(
    model=model_beta_gam,
    predictor=pred,
    model_label="Detected richness",
    ref_data=alpha.mod
  )
  
  pred_scaled_edna_list[[pred]]=predict_linear_gam_sitewise(
    model=model_beta_reads_gam,
    predictor=pred,
    model_label="eDNA index",
    ref_data=phylum_counts
  )
}

# Season predictions
pred_season_rich_gam=predict_categorical_gam_sitewise(
  model=model_beta_gam,
  predictor="SEASON",
  model_label="Detected richness",
  ref_data=alpha.mod
)

pred_season_edna_gam=predict_categorical_gam_sitewise(
  model=model_beta_reads_gam,
  predictor="SEASON",
  model_label="eDNA index",
  ref_data=phylum_counts
)
```

#### c. Plot 

```{r}
# For Richness (site-specific)
all_richness_fits_site=c()

# Collect all Richness predictions from all site-specific predictors
for(pred in scaled_predictors) {
  all_richness_fits_site=c(all_richness_fits_site, 
                              pred_scaled_rich_list[[pred]]$fit,
                              pred_scaled_rich_list[[pred]]$lower,
                              pred_scaled_rich_list[[pred]]$upper)
}

# Add cyclic predictors
all_richness_fits_site=c(all_richness_fits_site,
                            pred_wdir_rich_gam$fit,
                            pred_wdir_rich_gam$lower,
                            pred_wdir_rich_gam$upper,
                            pred_mwd_rich_gam$fit,
                            pred_mwd_rich_gam$lower,
                            pred_mwd_rich_gam$upper,
                            pred_lunar_rich_gam$fit,
                            pred_lunar_rich_gam$lower,
                            pred_lunar_rich_gam$upper)

# Add season
all_richness_fits_site=c(all_richness_fits_site,
                            pred_season_rich_gam$fit,
                            pred_season_rich_gam$lower,
                            pred_season_rich_gam$upper)

# Calculate global Richness range
global_richness_range_site=range(all_richness_fits_site, na.rm=TRUE)
global_richness_padding_site=diff(global_richness_range_site) * 0.15
global_richness_ylim_site=c(max(0, global_richness_range_site[1] - global_richness_padding_site), 
                               min(1, global_richness_range_site[2] + global_richness_padding_site))

# For eDNA Index (site-specific)
all_edna_fits_site=c()

for(pred in scaled_predictors) {
  all_edna_fits_site=c(all_edna_fits_site, 
                          pred_scaled_edna_list[[pred]]$fit,
                          pred_scaled_edna_list[[pred]]$lower,
                          pred_scaled_edna_list[[pred]]$upper)
}

# Add cyclic predictors
all_edna_fits_site=c(all_edna_fits_site,
                        pred_wdir_edna_gam$fit,
                        pred_wdir_edna_gam$lower,
                        pred_wdir_edna_gam$upper,
                        pred_mwd_edna_gam$fit,
                        pred_mwd_edna_gam$lower,
                        pred_mwd_edna_gam$upper,
                        pred_lunar_edna_gam$fit,
                        pred_lunar_edna_gam$lower,
                        pred_lunar_edna_gam$upper)

# Add season
all_edna_fits_site=c(all_edna_fits_site,
                        pred_season_edna_gam$fit,
                        pred_season_edna_gam$lower,
                        pred_season_edna_gam$upper)

# Calculate global eDNA range
global_edna_range_site=range(all_edna_fits_site, na.rm=TRUE)
global_edna_padding_site=diff(global_edna_range_site) * 0.15
global_edna_ylim_site=c(max(0, global_edna_range_site[1] - global_edna_padding_site), 
                           min(1, global_edna_range_site[2] + global_edna_padding_site))

cat("Global Richness range (site-specific):", round(global_richness_ylim_site, 4), "\n")
cat("Global eDNA Index range (site-specific):", round(global_edna_ylim_site, 4), "\n")



# Function to create cyclic plots with site-specific colors and GLOBAL SCALES
create_cyclic_site_plot=function(pred_data, predictor_name, x_label, 
                                    site_colors=cols.site,
                                    global_ylim=global_richness_ylim_site) {
  
  # Ensure SITE_site column exists
  if(!"SITE_site" %in% names(pred_data)) {
    pred_data$SITE_site=pred_data$SITE
  }
  
  plot=ggplot() +
    # Confidence ribbons for each site (semi-transparent)
    geom_ribbon(data=pred_data,
                aes(x=direction_deg, ymin=lower, ymax=upper, 
                    fill=SITE_site, group=SITE_site),
                alpha=0.15) +
    # Lines for each site
    geom_line(data=pred_data,
              aes(x=direction_deg, y=fit, color=SITE_site, group=SITE_site),
              size=1) +
    scale_color_manual(values=site_colors, name="Site") +
    scale_fill_manual(values=site_colors, name="Site") +
    scale_y_continuous(name="Predicted response", 
                       limits=global_ylim,
                       labels=scales::number_format(accuracy=0.01)) +
    labs(x=x_label, y="Predicted response") +
    My_Theme +
    theme(legend.position="right",
          legend.key.size=unit(0.3, "cm"))
  
  return(plot)
}

# Function to create scaled predictor plots with site-specific colors and GLOBAL SCALES
create_scaled_site_plot=function(pred_data, predictor_name, x_label,
                                    site_colors=cols.site,
                                    global_ylim=global_richness_ylim_site) {
  
  # Get the original scale column
  orig_col=paste0(predictor_name, "_orig")
  
  # Ensure SITE_site column exists
  if(!"SITE_site" %in% names(pred_data)) {
    pred_data$SITE_site=pred_data$SITE
  }
  
  plot=ggplot() +
    # Confidence ribbons for each site
    geom_ribbon(data=pred_data,
                aes(x=.data[[orig_col]], ymin=lower, ymax=upper, 
                    fill=SITE_site, group=SITE_site),
                alpha=0.15) +
    # Lines for each site
    geom_line(data=pred_data,
              aes(x=.data[[orig_col]], y=fit, color=SITE_site, group=SITE_site),
              size=1) +
    scale_color_manual(values=site_colors, name="Site") +
    scale_fill_manual(values=site_colors, name="Site") +
    scale_y_continuous(name="Predicted response", 
                       limits=global_ylim,
                       labels=scales::number_format(accuracy=0.01)) +
    labs(x=x_label, y="Predicted response") +
    My_Theme +
    theme(legend.position="right",
          legend.key.size=unit(0.3, "cm"))
  
  return(plot)
}

# Function for categorical predictor with GLOBAL SCALES
create_season_site_plot=function(pred_data, site_colors=cols.site.paired,
                                    global_ylim=global_richness_ylim_site) {
  
  plot=ggplot(pred_data) +
    geom_point(aes(x=SEASON_val, y=fit, color=SITE_site),
               position=position_dodge(width=0.5), size=2) +
    geom_errorbar(aes(x=SEASON_val, ymin=lower, ymax=upper, color=SITE_site),
                  position=position_dodge(width=0.5), width=0.2) +
    scale_color_manual(values=site_colors, name="Site") +
    scale_y_continuous(name="Predicted response", 
                       limits=global_ylim,
                       labels=scales::number_format(accuracy=0.01)) +
    labs(x="Season", y="Predicted response") +
    My_Theme +
    theme(legend.position="right",
          legend.key.size=unit(0.3, "cm"))
  
  return(plot)
}


# Cyclic plots - Richness (using global richness scale)
cyclic_wdir_rich=create_cyclic_site_plot(
  pred_wdir_rich_gam, "WDIR", "Wind direction (°)", 
  cols.site, global_richness_ylim_site)

cyclic_mwd_rich=create_cyclic_site_plot(
  pred_mwd_rich_gam, "MWD", "Wave direction (°)", 
  cols.site, global_richness_ylim_site)

cyclic_lunar_rich=create_cyclic_site_plot(
  pred_lunar_rich_gam, "lunar", "Lunar angle (°)", 
  cols.site, global_richness_ylim_site)

# Cyclic plots - eDNA Index (using global eDNA scale)
cyclic_wdir_edna=create_cyclic_site_plot(
  pred_wdir_edna_gam, "WDIR", "Wind direction (°)", 
  cols.site, global_edna_ylim_site)

cyclic_mwd_edna=create_cyclic_site_plot(
  pred_mwd_edna_gam, "MWD", "Wave direction (°)", 
  cols.site, global_edna_ylim_site)

cyclic_lunar_edna=create_cyclic_site_plot(
  pred_lunar_edna_gam, "lunar", "Lunar angle (°)", 
  cols.site, global_edna_ylim_site)

# Scaled predictor plots - Richness
scaled_sst_rich=create_scaled_site_plot(
  pred_scaled_rich_list[["SST_scaled"]], "SST_scaled", "SST (°C)", 
  cols.site, global_richness_ylim_site)

scaled_wind_rich=create_scaled_site_plot(
  pred_scaled_rich_list[["wind_scaled"]], "wind_scaled", "Wind speed (m/s)", 
  cols.site, global_richness_ylim_site)

scaled_wave_rich=create_scaled_site_plot(
  pred_scaled_rich_list[["wave_scaled"]], "wave_scaled", "Wave height (m)", 
  cols.site, global_richness_ylim_site)

scaled_tide_rich=create_scaled_site_plot(
  pred_scaled_rich_list[["tide_scaled"]], "tide_scaled", "Tide height (m)", 
  cols.site, global_richness_ylim_site)

# Scaled predictor plots - eDNA Index
scaled_sst_edna=create_scaled_site_plot(
  pred_scaled_edna_list[["SST_scaled"]], "SST_scaled", "SST (°C)", 
  cols.site, global_edna_ylim_site)

scaled_wind_edna=create_scaled_site_plot(
  pred_scaled_edna_list[["wind_scaled"]], "wind_scaled", "Wind speed (m/s)", 
  cols.site, global_edna_ylim_site)

scaled_wave_edna=create_scaled_site_plot(
  pred_scaled_edna_list[["wave_scaled"]], "wave_scaled", "Wave height (m)", 
  cols.site, global_edna_ylim_site)

scaled_tide_edna=create_scaled_site_plot(
  pred_scaled_edna_list[["tide_scaled"]], "tide_scaled", "Tide height (m)", 
  cols.site, global_edna_ylim_site)

# Season plots
season_rich=create_season_site_plot(
  pred_season_rich_gam, cols.site.paired, global_richness_ylim_site)

season_edna=create_season_site_plot(
  pred_season_edna_gam, cols.site.paired, global_edna_ylim_site)


richness_plots_no_legend=list(
  scaled_wind_rich + theme(legend.position="none"),
  cyclic_wdir_rich + theme(legend.position="none"),
  scaled_wave_rich + theme(legend.position="none"),
  cyclic_mwd_rich + theme(legend.position="none"),
  scaled_tide_rich + theme(legend.position="none"),
  cyclic_lunar_rich + theme(legend.position="none"),
  scaled_sst_rich + theme(legend.position="none"),
  season_rich + theme(legend.position="none")
)

edna_plots_no_legend=list(
  scaled_wind_edna + theme(legend.position="none"),
  cyclic_wdir_edna + theme(legend.position="none"),
  scaled_wave_edna + theme(legend.position="none"),
  cyclic_mwd_edna + theme(legend.position="none"),
  scaled_tide_edna + theme(legend.position="none"),
  cyclic_lunar_edna + theme(legend.position="none"),
  scaled_sst_edna + theme(legend.position="none"),
  season_edna + theme(legend.position="none")
)

# Extract the legend from the first plot
legend=get_legend(scaled_wind_rich + theme(legend.position="bottom"))

# Row 1: All Richness plots (8 columns)
row_richness=ggarrange(
  plotlist=richness_plots_no_legend,
  ncol=8, nrow=1,
  labels=c("a", "b", "c", "d", "e", "f", "g", "h"),
  font.label=list(size=12, face="bold")
)

# Row 2: All eDNA plots (8 columns)
row_edna=ggarrange(
  plotlist=edna_plots_no_legend,
  ncol=8, nrow=1,
  labels=c("i", "j", "k", "l", "m", "n", "o", "p"),
  font.label=list(size=12, face="bold")
)

# Combine rows with a shared legend
figure_site_specific_global=ggarrange(
  row_richness, row_edna,
  ncol=1, nrow=2,
  common.legend=TRUE,
  legend="bottom",
  legend.grob=legend,
  heights=c(1, 1)
)

# Add labels for response types on the left
GAM_site=annotate_figure(
  figure_site_specific_global,
  left=text_grob("Richness", rot=90, size=14, face="bold", 
                   color="#1985a1", vjust=0.5, y=0.75),
  bottom=text_grob("eDNA Index", size=14, face="bold", 
                     color="black", vjust=0.5)
)

print(GAM_site)

ggsave(filename=file.path(paths$results_figs, "FIGURE_4_GAM_site.pdf"), plot=GAM_site, width=22, height=8, units="in", device=cairo_pdf, family="Arial")
```


### v. Observed frequency histograms

```{r}
create_frequency_histogram=function(data, predictor, x_label, 
                                       title_label, color_fill="tan",
                                       bins=30, y_max=1) {
  
  # Extract the variable
  var_data=data[[predictor]]
  
  # Remove NAs
  var_data=var_data[!is.na(var_data)]
  
  # Create data frame for plotting
  plot_data=data.frame(value=var_data)
  
  # Create histogram with scaled frequency (max=1)
  p=ggplot(plot_data, aes(x=value)) +
    geom_histogram(
      aes(y=after_stat(count / max(count))),
      bins=bins,
      fill=color_fill,
      alpha=0.3,
      color=NA,
      size=0.3
    ) +
    scale_y_continuous(
      name="Relative frequency",
      limits=c(0, y_max),
      breaks=seq(0, y_max, by=0.25)
    ) +
    scale_x_continuous(
      name=x_label
    ) +
    labs(title=title_label) +
    My_Theme +
    theme(
      plot.title=element_text(size=12, face="bold", hjust=0.5),
      axis.title=element_text(size=10),
      axis.text=element_text(size=8)
    )
  
  return(p)
}


# 1. Wind Speed (WSPD)
hist_wind=create_frequency_histogram(
  data=alpha.mod,
  predictor="WSPD",
  x_label="Wind speed (m/s)",
  title_label="A",
  color_fill="tan"
)

# 2. Wind Direction (WDIR)
hist_wdir=create_frequency_histogram(
  data=alpha.mod,
  predictor="WDIR",
  x_label="Wind direction (°)",
  title_label="B",
  color_fill="tan"
)

# 3. Wave Height (WVHT)
hist_wave=create_frequency_histogram(
  data=alpha.mod,
  predictor="WVHT",
  x_label="Wave height (m)",
  title_label="C",
  color_fill="tan"
)

# 4. Wave Direction (MWD)
hist_mwd=create_frequency_histogram(
  data=alpha.mod,
  predictor="MWD",
  x_label="Wave direction (°)",
  title_label="D",
  color_fill="tan"
)

# 5. Tide Height
hist_tide=create_frequency_histogram(
  data=alpha.mod,
  predictor="tide_height_m",
  x_label="Tide height (m)",
  title_label="E",
  color_fill="tan"
)

# 6. Lunar Angle (degrees)
hist_lunar=create_frequency_histogram(
  data=alpha.mod,
  predictor="degrees",
  x_label="Lunar angle (°)",
  title_label="F",
  color_fill="tan"
)

# 7. Sea Surface Temperature (SST)
hist_sst=create_frequency_histogram(
  data=alpha.mod,
  predictor="SST",
  x_label="SST (°C)",
  title_label="G",
  color_fill="tan"
)

# 8. Season (categorical - special handling)
season_data=data.frame(
  SEASON=alpha.mod$SEASON
)

season_counts=season_data %>%
  group_by(SEASON) %>%
  summarise(count=n()) %>%
  mutate(rel_freq=count / max(count))

hist_season=ggplot(season_counts, aes(x=SEASON, y=rel_freq)) +
  geom_bar(stat="identity", fill="tan", alpha=0.3, color=NA, size=0.3) +
  scale_y_continuous(
    name="Relative frequency",
    limits=c(0, 1),
    breaks=seq(0, 1, by=0.25)
  ) +
  scale_x_discrete(
    name="Season"
  ) +
  labs(title="H") +
  My_Theme +
  theme(
    plot.title=element_text(size=12, face="bold", hjust=0.5),
    axis.title=element_text(size=10),
    axis.text=element_text(size=8)
  )

figure_histograms=ggarrange(
  hist_wind, hist_wdir,
  hist_wave, hist_mwd,
  hist_tide, hist_lunar,
  hist_sst, hist_season,
  ncol=4, nrow=2,
  labels=NULL  # Labels are already in the plots
)

# Add overall title
figure_histograms=annotate_figure(
  figure_histograms,
  top=text_grob("Observed Frequency Distributions of Environmental Predictors", 
                  size=16, face="bold")
)

# Print the figure
print(figure_histograms)

ggsave(filename=file.path(paths$results_figs, "FIGURE_4_histograms_only.pdf"), plot=figure_histograms, width=12, height=6, units="in", device=cairo_pdf, family="Arial")
```

##F. Figure 6

### i. Prepare eDNA index timeseries data

```{r}
data.long.top5=subset(data.long.top5, YEAR != "2023")
data.long.top5=subset(data.long.top5, STEP != "1")
data.long.top5$DATE=as.Date(data.long.top5$DATE, format="%d/%m/%Y")

Modelplot.step.prop=data.long.top5 %>%
  group_by(STEP, DATE, SITE, MARKER, phylum) %>%
  summarise(
    index=sum(value, na.rm=TRUE),
    sd=sd(value, na.rm=TRUE),
    se=sd / sqrt(n()),
    .groups="drop"
  ) %>%
  group_by(SITE, MARKER, STEP) %>%
  mutate(total_value=sum(index, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(
    prop=index / total_value,
    DATE=as.Date(DATE, format="%d/%m/%Y")
  ) %>%
  arrange(SITE, phylum, STEP, MARKER)

Modelplot.step.index=data.long.top5 %>%
  group_by(SITE, phylum, STEP, DATE, MARKER) %>%
  summarise(
    index=sum(value, na.rm=TRUE),
    .groups="drop"
  ) %>%
  mutate(
    STEP=as.numeric(STEP),
    DATE=as.Date(DATE, format="%d/%m/%Y")
  )

final_index=Modelplot.step.index %>%
  select(SITE, phylum, STEP, MARKER, index) %>%
  pivot_wider(
    names_from=MARKER,
    values_from=index
  ) %>%
  filter(!is.na(`18S`), !is.na(COI)) %>%
  arrange(SITE, phylum, STEP) %>%
  distinct(SITE, phylum, STEP, .keep_all=TRUE)

```

### ii. Calculate concordance

```{r}
# Compute Kendall W for paired observations among SITE x Phylum x Marker combinations
library(synchrony)
set.seed(123)
sync_results=final_index %>%
  group_by(SITE, phylum) %>%
  summarise(
    n_steps=n(),
    ts=list(
      as.matrix(pick(`18S`, COI)[order(STEP), ])
    ),
    .groups="drop"
  ) %>%
  filter(n_steps >= 6) %>%
  rowwise() %>%
  mutate(
    sync_res=list(
      kendall.w(
        data  =ts,
        nrands=numperm,
        type  =1,   # or 2
        quiet =FALSE
      )
    ),
    W      =sync_res$w.corrected,
    rho    =sync_res$spearman.corr,
    p_value=sync_res$pval.rand
  ) %>%
  ungroup() %>%
  select(SITE, phylum, n_steps, W, rho, p_value)


annot_data=sync_results %>%
  left_join(
    Modelplot.step.prop %>% distinct(SITE, phylum, DATE),
    by=c("SITE", "phylum")
  ) %>%
  mutate(
    sync_xpos=min(DATE) - 5,
    r_xpos   =min(DATE) + 225,
    p_xpos   =min(DATE) + 460,
    sync_ypos=0.95,
    y_pos_r  =0.95,
    y_pos_p  =0.95
  ) %>%
  distinct(SITE, phylum, .keep_all=TRUE)
```

### iii. Plot

```{r}
fig6=ggplot(Modelplot.step.prop, aes(x=DATE, y=prop, group=interaction(SITE, MARKER), color=MARKER)) + 
  geom_vline(xintercept=as.Date(c("2019-09-19", "2019-11-27", "2020-09-22", "2020-11-30")), 
             linetype="dashed", color="grey70", size=0.7, alpha=0.6) +
  geom_line() +
  geom_point(size=1) +
  geom_errorbar(aes(ymin=prop - se, ymax=prop + se), width=0.2) +
  xlab("") +
  ylab("Relative eDNA Contribution") + 
  #scale_y_sqrt(limits=c(0, 1)) +
  ylim(0,1)+
  facet_grid(SITE ~ phylum, scales="fixed") + 
  scale_color_manual(values=cols.marker) + 
  scale_x_date(
  breaks="4 months",
  labels=scales::date_format("%m/%y"),
  expand=c(.1, .1)
) + 
  My_Theme +
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=0.5),
    axis.text.y=element_text(size=10),
    panel.spacing=unit(1, "lines"), 
    legend.position="right"
  ) +
  # Add Kendall's W annotation (top-left, slightly below max)
  geom_text(
    aes(
      x=sync_xpos, y=sync_ypos,  
      label=paste0("W= ", formatC(W, format="f", digits=2)),
      fontface=ifelse(p_value < 0.05, "bold", "plain")  # Bold if p-value is significant
    ),
    data=annot_data,
    inherit.aes=FALSE,
    size=3.2,
    hjust=0  # Align left
  ) +
  geom_text(aes(x=r_xpos, y=y_pos_r,label=paste0("\u03C1= ", formatC(rho, format="f", digits=2)), fontface=ifelse(p_value < 0.05, "bold", "plain")),data=annot_data,inherit.aes=FALSE,size=3.2, hjust=0) +
  # Add p-value annotation
  geom_text(
    aes(
      x=p_xpos, y=y_pos_p, 
      label=ifelse(p_value < 0.01, "p < 0.01", paste0("p=", formatC(p_value, format="f", digits=2))),
      fontface=ifelse(p_value < 0.05, "bold", "plain")  # Bold if p-value is significant
    ),
    data=annot_data,
    inherit.aes=FALSE,
    size=3.2,
    hjust=0  # Align left
  ) + theme(legend.position="none")
fig6
```

###iv. Export

```{r}
ggsave(
  filename=file.path(paths$results_figs, "FIGURE_6.pdf"),
  plot    =fig6,
  width   =11,
  height  =6,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


## G. RF Models

###i. Richness

```{r}
alpha.mod.clean=alpha.mod %>%
  drop_na(wave_scaled, MWD) #removing missing wave data observations

#### RF Model ###################################################
rf_perm_prop=rfPermute(formula=prop ~ SST_scaled + wave_scaled + tide_scaled + wind_scaled + MWD + WDIR + degrees + SEASON + MARKER+ SITE, data=alpha.mod.clean, ntree=250, num.rep=numperm, num.cores=cores_to_use)
saveRDS(rf_perm_prop, file.path(paths$results_rds,"RFmodel_richness.rds"))
##############################################################
```


###ii. Reads

```{r}
#### RF Model ################################
rf_perm_reads=rfPermute(formula=reads ~ SST_scaled + wave_scaled + tide_scaled + wind_scaled + MWD + WDIR + degrees + SEASON + phylum + MARKER+ SITE, data=phylum_counts_clean, ntree=100, num.rep=numperm, num.cores=cores_to_use)
saveRDS(rf_perm_reads, file.path(paths$results_rds,"RFmodel_reads.rds"))
############################################
```


### iii. Dominant Phyla

```{r}
phylum_counts_pa=phylum_counts_clean %>%
  mutate(present=ifelse(abundance > 0, "Present", "Absent"))

## Dominant Phyla
dominant_phylum=phylum_counts_pa %>%
  group_by(SAMPLE) %>%
  slice_max(abundance, n=1) %>%
  ungroup()

####RF Model ###############################################
rf_perm_phyla=rfPermute(as.factor(phylum) ~ SST_scaled + wave_scaled + tide_scaled + wind_scaled + MWD_sin + MWD_cos + WDIR_sin + WDIR_cos + lunar_sin + lunar_cos + SEASON + MARKER+ SITE, data=dominant_phylum, ntree=500, importance=TRUE, sampsize=rep(min(table(dominant_phylum$phylum)), length(unique(dominant_phylum$phylum))), num.rep=numperm, num.cores=cores_to_use)
saveRDS(rf_perm_phyla, file.path(paths$results_rds,"RFmodel_phyla.rds"))
##############################################################
rf_perm_phyla=readRDS(file.path(paths$results_rds,"RFmodel_phyla.rds"))

rf_perm_phyla
plotImportance(rf_perm_phyla)

oob.error=as.data.frame(rf_perm_phyla$rf$confusion)
oob.error$accuracy=1-oob.error$class.error
oob.error$mean_err=mean(rf_perm_phyla$rf$confusion[,6])#OOB Accuracy
oob.error$mean_accur=1-mean(rf_perm_phyla$rf$confusion[,6]) #OOB Accuracy
write.csv(oob.error, file.path(paths$results_tables, "TABLE_RFclassification_error.csv"), row.names=FALSE)
```
